---
title: "UAS Pointclouds vs. Lidar Pointclouds for structual analysis of forests"
output: github_document
bibliography: /home/marvin/literature/zotero/references.bib
---

```{r setup, include = FALSE}
source("../scripts/000_setup.R")
library(ggpubr)


ggHexCor = function(df, l = "lidar_2018_01_01", d){
  ggplot(df, aes_(x = as.name(l), y = as.name(d)))+
    geom_hex()+
    scale_fill_gradientn(name = "Count",colors = viridis(25))+
    geom_abline(slope = 1, intercept = 0, color = "red", lwd = 1)+
    facet_wrap(.~metric, scales = "free")+
    theme_bw()
}


fls = list.files("../data/z_metrics_tables/", pattern = ".RDS$", full.names = TRUE)
df = do.call(rbind, lapply(fls, readRDS))

df = df %>% filter(metric != "n_points" & metric != "z_shannon")
```

# Introduction


The use of Light detection and ranging (LiDAR) pointclouds are well established in forestry, agriculture and forest research. Spatially extensive estimations of vegetation related structural parameters are mostly realised with the calculation of LiDAR indices in a regular grid. Common applications on a forest stand scale are the retrieval of canopy cover and heights [@Lee2018; @Alexander2014], stand density [@Lee2018a] and the estimation of leaf area [@Kamoske2019]. The structural information and its heterogeneity in a landscape serve as indicators for biodiversity [@Hilmers2018] or species occurrence [@Carrasco2019; @Melin2016; @Froidevaux2016]. Most of these Lidar indices are strongly correlated [@Shi2018].

Despite their relevance in forest applications, Lidar data has some major drawbacks, mainly in their cost and accessibility. Lidar sensors and data acquisition are expensive and often distributed commercially. Data provided by governmental institutions are for the most part still irregularly available and not publicly distributed. Further, the temporal resolution of the data is low (by law every 3 years in Germany) making them not suitable for monitoring or applications which require different seasonal conditions. With the recent development of unmanned aerial systems (UAS) and photogrammetric techniques like structure from motion (SfM), an alternative to Lidar pointclouds are available. Quick data access in moderately large areas makes UAS data promising for the monitoring of agricultural or natural systems [@Manfreda2018]. Depending on flight conditions, these pointclouds could be acquired on a near daily basis. Especially in forest environments, research can benefit from vegetation structural information retrieved from UAS data.

UAS pointclouds do not have return values which many Lidar indices depend on. Every point is a first return so we cannot get below a developed canopy. These return values however are crucial for LiDAR point classification (e.g. differentiate between ground and non ground point).

The quality and viability of UAS pointclouds have to be assessed in terms of comparability to Lidar pointclouds (since Lidar structural analysis is the standard in many studies)

## Epic 1: Similarities between Lidar and UAS pointclouds


Since photogrammetically received pointclouds only capture the surface and do not penetrate the forest canopy like Lidar pointclouds, different phenological stages should capture different vertical layers of the forest canopy. Therefore, the photogrammetrically received pointcloud should represent and correlate with different parts of the Lidar pointcloud. E.g. a flight without leafs in winter or early spring should correlate well with the last returns of a lidar pointcloud (or a leaf-off lidar campaign) and therefore should be suitable for creating elevation models or the detection of tree stems and branches.

Further, a UAS flight at times with a fully developed canopy leads to pointclouds where it is very unlikely to capture ground points. These pointclouds should be comparable to first returns of a lidar pointcloud. E.g. it was previously shown, that UAS pointclouds are very promising for the estimation of tree heights and canopy cover.

**Hypothesis 1:** Photogrammetrically received pointclouds from different phenological stages in a deciduous forest correlate with different parts of a LiDAR derived pointcloud.

This relationship will be shown by comparing the vertical distributions of both pointclouds in a regular grid. Further, the Lidar data will be filtered to different return counts in order to check, which part of the lidar data is represented.

## Epic 2: Multitemporal UAS pointclouds compared to Lidar pointclouds


If the positional accuracy of the individual photogrammetric pointclouds is high enough (previously shown in @Ludwig2020, it is a resonable assumption to combine pointclouds from different phenological stages in order to get a full 3D model of the forest. The positional accuracy can be validated with tree positions and stem axis surveyed with a totalstation **(How?)**.

**Hypothesis 2:** Mutlitemporal photogrammetrically received pointclouds can substitude Lidar derived pointclouds for forest structural analysis.

This relation will be shown by comparing commonly used structural indices like the `penetration rate` or the estimation of `biomass` from both pointcloud types. Further, the structural data will be validated by field work around at 15 plots where the vertical structure of the forest was assessed.

# Methods

## Study Area


The study area is a 200 x 150 m part of a mixed deciduous forest near Marburg (Germany). The area consists of a mix of oaks (\\textit{Quercus spec.}) and beeches (\\textit{Fagus sylvatica}) and represent a typical environment in a managed forest. The elevation ranges from XXXm to XXXm a.s.l. Stem positions of 500 trees were acquired by using a differential GPS (Zenith 35 Pro, GeoMax Widnau Switzerland) with a positioning accuracy of 0.05 m.\\

# Comparing height based metrics

## Lidar vs. Densecloud 2019-04-25

```{r, warning=FALSE, echo = FALSE}
ggHexCor(df, d = "densecloud_2019_04_25")
```

## Lidar vs. Densecloud 2019-12-04

```{r, warning=FALSE, echo = FALSE}
ggHexCor(df, d = "densecloud_2019_12_04")
```


## Lidar vs. Densecloud 2020-09-15

```{r, warning=FALSE, echo = FALSE}
ggHexCor(df, d = "densecloud_2020_09_15")
```


## Lidar vs. Densecloud 2020-11-12

```{r, warning=FALSE, echo = FALSE}
ggHexCor(df, d = "densecloud_2020_11_12")
```

# Using only lidar first returns

Indices used in [@Carrasco2019]


## Lidar Firstreturns vs. Densecloud 2019-04-25

```{r, warning=FALSE, echo = FALSE}
ggHexCor(df, l = "lidarfirst_2018_01_01",d = "densecloud_2019_04_25")
```

## Lidar Firstreturns vs. Densecloud 2019-12-04

```{r, warning=FALSE, echo = FALSE}
ggHexCor(df, l = "lidarfirst_2018_01_01", d = "densecloud_2019_12_04")
```


## Lidar Firstreturns vs. Densecloud 2020-09-15

```{r, warning=FALSE, echo = FALSE}
ggHexCor(df, l = "lidarfirst_2018_01_01",d = "densecloud_2020_09_15")
```


## Lidar Firstreturns vs. Densecloud 2020-11-12

```{r, warning=FALSE, echo = FALSE}
ggHexCor(df, l = "lidarfirst_2018_01_01",d = "densecloud_2020_11_12")
```




# Text fragments


The main challange for further usage of Lidar data in a forest environment is the detection of individual trees. This enables the estimation of tree related parameters such as diameter at breast height, timber volume or crown related metrics [@Leeuwen2010]. Forest structure then can be described as the sum of the structure of individual trees [e.g. their height and biomass @Ferraz2016] and the species composition (REF). This could give new insights into ecosystem functioning, since many processes and species distributions depend on functions provided by trees or their related microhabitats (REF). Further, monitoring of individual tree health and drought could be applied in forestry.


# References

